---
title: "Getting Started with ggsynteny"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ggsynteny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

## Introduction

`ggsynteny` provides tools for creating publication-quality synteny plots for comparative genomics. This vignette demonstrates the basic usage of the package.

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("yourusername/ggsynteny")
```

```{r}
library(ggsynteny)
library(ggplot2)
```

## Macro-Synteny Plots

### Basic Example

Let's start with the bundled example data showing synteny between Arabidopsis, Grape, and Rice:

```{r}
# Load example data
syn <- example_synteny_data()

# Inspect structure
str(syn)
```

The data contains two data frames:
- `chromosomes`: chromosome sizes for each species
- `blocks`: syntenic block coordinates

### Creating a Basic Plot

```{r fig.height=8}
# Define species order (top to bottom)
sp_order <- c("Arabidopsis", "Grape", "Rice")

# Create plot
p1 <- plot_synteny(syn,
                   species_order = sp_order,
                   title = "Plant Genome Synteny")
print(p1)
```

### Using the Built-in Palettes

Every `palette` argument accepts, by name, all 32 palettes of the
[ltc package](https://github.com/loukesio/ltc-color-palettes) (vendored, so
ltc need not be installed). One palette can drive the whole plot:

```{r fig.height=8}
p_pal <- plot_synteny(syn, sp_order, palette = "casa_natal")
print(p_pal)
```

List them all with `syn_palettes()`; names match case-insensitively and
ignore spaces, underscores, and dashes:

```{r}
names(syn_palettes())
```

### Customizing Colors

#### Per-Species Chromosome Colors

```{r fig.height=8}
p2 <- plot_synteny(syn, sp_order,
                   chr_fill = "per_species",
                   chr_palette = c(
                     "Arabidopsis" = "#B8D4E3",
                     "Grape"       = "#C5B4E3",
                     "Rice"        = "#B8E3C5"
                   ),
                   ribbon_fill = "uniform",
                   ribbon_palette = "#50B88E",
                   ribbon_alpha = 0.25,
                   title = "Custom Species Colors")
print(p2)
```

#### Per-Chromosome Colors

```{r fig.height=8}
# Define color palette for chromosomes 1-12
chr_pal <- c("1"="#1B3A8C", "2"="#3B7DD8", "3"="#00BCD4", "4"="#4CAF50",
             "5"="#8BC34A", "6"="#CDDC39", "7"="#9C27B0", "8"="#E91E63",
             "9"="#FF5722", "10"="#FF9800", "11"="#795548", "12"="#F44336")

p3 <- plot_synteny(syn, sp_order,
                   chr_fill = "per_chr",
                   chr_palette = chr_pal,
                   ribbon_fill = "source_chr",
                   ribbon_palette = chr_pal,
                   title = "Per-Chromosome Coloring")
print(p3)
```

### Species-Pair Ribbon Colors

```{r fig.height=8}
p4 <- plot_synteny(syn, sp_order,
                   chr_fill = "uniform",
                   chr_palette = "#E8E4DF",
                   ribbon_fill = "species_pair",
                   ribbon_palette = c(
                     "Arabidopsis_Grape" = "#3A6EA5",
                     "Grape_Rice"        = "#E8573A"
                   ),
                   title = "Species-Pair Ribbon Colors")
print(p4)
```

## Micro-Synteny Plots

### Basic Example

Micro-synteny plots show gene-level conservation:

```{r fig.height=6}
# Load microsynteny example
micro <- demo_microsynteny_data()

# Inspect structure
str(micro)
```

The data contains:
- `features`: gene coordinates, strand, and names
- `links`: homology relationships with identity scores

### Creating a Gene-Level Plot

```{r fig.height=6}
p5 <- plot_microsynteny(micro$features,
                        micro$links,
                        bin_order = c("ZONMW-30", "ZONMW-20", "ZONMW-10"),
                        gene_fill = "per_name",
                        ribbon_fill = "identity",
                        title = "Micro-synteny: moa/moe Cluster")
print(p5)
```

### Custom Gene Colors

```{r fig.height=6}
# Define custom gene colors
gene_pal <- c(
  moaE   = "#E06C75",
  moaC2  = "#61AFEF",
  moaA   = "#98C379",
  moeA   = "#E5C07B",
  mobA   = "#C678DD",
  spacer = "#DDDDDD"
)

p6 <- plot_microsynteny(micro$features,
                        micro$links,
                        bin_order = c("ZONMW-30", "ZONMW-20", "ZONMW-10"),
                        gene_fill = "per_name",
                        gene_palette = gene_pal,
                        ribbon_fill = "identity",
                        ribbon_alpha = 0.40,
                        title = "Custom Gene Colors")
print(p6)
```

## Reading Your Own Data

### TSV Format

```{r eval=FALSE}
# Read your own TSV files
syn <- read_synteny_tsv("my_chromosomes.tsv", "my_blocks.tsv")
p <- plot_synteny(syn, species_order = c("SpeciesA", "SpeciesB", "SpeciesC"))
```

### MCScanX Format

```{r eval=FALSE}
# Parse MCScanX output
syn <- read_mcscanx("output.collinearity", "output.gff")
p <- plot_synteny(syn, species_order = c("Species1", "Species2"))
```

### GENESPACE Format

```{r eval=FALSE}
# Parse GENESPACE output
syn <- read_genespace("synHits.tsv")
p <- plot_synteny(syn, species_order = c("genome1", "genome2", "genome3"))
```

## Saving Plots

```{r eval=FALSE}
# Save as high-resolution PNG
ggsave("synteny.png", p1, width = 20, height = 10, dpi = 300, bg = "white")

# Save as PDF for publications
ggsave("synteny.pdf", p1, width = 20, height = 10)

# Save as SVG for editing in Inkscape/Illustrator
ggsave("synteny.svg", p1, width = 20, height = 10)
```

## Advanced Customization

Since `ggsynteny` returns ggplot2 objects, you can further customize them:

```{r fig.height=8}
p7 <- plot_synteny(syn, sp_order,
                   chr_fill = "per_species",
                   ribbon_fill = "source_chr") +
  theme(plot.background = element_rect(fill = "white", color = NA)) +
  labs(caption = "Data source: Example synteny data")

print(p7)
```

## Session Info

```{r}
sessionInfo()
```
