---
title: "GC content and annotation tracks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GC content and annotation tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 7, warning = FALSE, message = FALSE)
library(ggsynteny)
```

Add a GC-content strip to any synteny plot with `p + syn_track(values)`.
The same table works in linear and circular views. Choose `geom = "heatmap"`,
`"line"`, or `"bar"`, and combine them with `+`. The underlying gene and ribbon
colors keep their own meaning. All drawing uses ggplot2 layers.

## A complete gene example

These bundled DNA strings are simulated. They demonstrate the calculation and
alignment; their GC values are not measurements of real organisms.

```{r input}
example_dir <- system.file("extdata", "gc-tracks", package = "ggsynteny")
read_example <- function(name) {
  readr::read_tsv(file.path(example_dir, paste0(name, ".tsv")), show_col_types = FALSE)
}
features <- read_example("features")
links <- read_example("links")
dna <- read_example("sequences")
gene_gc <- gc_content(dna, intervals = features)
```

```{r linear}
p <- plot_microsynteny(features, links, palette = "casa_natal",
                      ribbon_fill = "per_name")
p + syn_track(gene_gc) + ggplot2::labs(title = "GC content per gene: simulated DNA")
```

```{r circular}
plot_circular_microsynteny(features, links, palette = "casa_natal",
                          ribbon_fill = "per_name") +
  syn_track(gene_gc) + ggplot2::labs(title = "The same GC values around a circle")
```

GC is the percentage of called bases that are G or C:
`100 * (G + C) / (A + C + G + T)`. Ambiguous bases such as `N` are left out of
the denominator. `n_called` and `n_ambiguous` let you assess how much sequence
supports each value. An interval with no called bases has `NA`, drawn grey.
Missing intervals remain gaps. Neither is treated as zero.

## Chromosome windows

```{r windows}
syn <- list(chromosomes = read_example("chromosomes"), blocks = read_example("blocks"))
window_gc <- gc_content(dna, window = 250)
plot_synteny(syn, unique(syn$chromosomes$species), palette = "casa_natal") +
  syn_track(window_gc, name = "GC (%) / 250 bp") +
  ggplot2::theme(plot.margin = ggplot2::margin(10, 30, 10, 100))
plot_circular_synteny(syn, palette = "casa_natal") +
  syn_track(window_gc, name = "GC (%) / 250 bp")
```

The final window is shortened at a sequence end. Set `step` smaller than
`window` to calculate sliding windows. Their intervals overlap; tiles draw in
input order, with later rows on top. For an unambiguous heatmap, use
non-overlapping windows or reduce overlapping measurements to disjoint bins
before plotting. A line track instead plots the value at each window center,
so overlapping sliding windows can be displayed directly. Intervals crossing a circular origin must first be split;
the calculator does not wrap windows around the origin.

## Supplying your own measurements

The common table has five columns:

| Column | Meaning |
| --- | --- |
| `group` | Species in chromosome plots; bin/genome in gene plots |
| `seq_id` | Chromosome or contig identifier within that group |
| `start`, `end` | Interval boundaries in the plot's coordinate system |
| `value` | Numeric measurement; GC uses percentages from 0 to 100 |

Existing `species`/`chr` and `bin_id`/`seq_id` columns also work. Gene feature
tables can be passed directly after adding a `value` column. Rows for groups
not shown in the plot are omitted with a warning. Unknown sequences within a
displayed group and intervals outside displayed bounds cause an error.
Gene plots only display each contig's first-to-last feature span; restrict
window tables to those bounds before adding them.

`gc_content()` reads character DNA strings, with sequence coordinate zero at
the first base. Its intervals use **zero-based, half-open base-pair coordinates**:
`start = 0, end = 4` selects bases 1 through 4. Convert one-based inclusive
gene annotations by subtracting 1 from `start` and leaving `end` unchanged.
If a chromosome plot uses kb or Mb, convert the calculated interval coordinates
to that unit before plotting. GC values cannot be obtained from synteny links,
block ranks, or gene positions alone.

For unavailable sequence, supply `NA_character_` in `sequence` and explicit
intervals. Missing sequence identifiers are errors. Sequences with no length
cannot be used to create windows. Lowercase and whitespace are accepted; gaps
and non-DNA characters are rejected. Reverse-complementing a gene does not
change its GC content.

## Multiple tracks and custom colors

The first addition is closest to the chromosome or gene row. Later additions
stack outward, with their own legends. Labels move outward with the tracks.

```{r mixed-types}
sliding_gc <- gc_content(dna, window = 400, step = 100)
ambiguous <- gc_content(dna, window = 100)
ambiguous$value <- 100 * ambiguous$n_ambiguous / (ambiguous$end - ambiguous$start)
tracks <- list(
  syn_track(gene_gc, geom = "heatmap", name = "Gene GC (%)", height = 0.055),
  syn_track(sliding_gc, geom = "line", name = "Window GC (%)",
            height = 0.19, gap = 0.025, colour = "#176D81", reference = 50),
  syn_track(ambiguous, geom = "bar", name = "Ambiguous bases (%)",
            height = 0.11, gap = 0.025, colour = "#BE7442")
)
plot_microsynteny(features, links, palette = "casa_natal", label_genes = FALSE) + tracks
plot_circular_microsynteny(features, links, palette = "casa_natal", label_genes = FALSE) + tracks
```

Each track owns its value range, set through `limits`. Lines and bars increase
upward in a linear plot and outward in a circular plot. The edge labels show
the lower and upper limits. `reference` adds a dashed guide at a chosen value.
Use `axis = FALSE` to hide the edge labels and `show.legend = FALSE` to hide
the track's legend. `colour` sets line/bar color; `linewidth` sets line thickness.
Heatmap colors still use `palette` and `scale_fill_syn_track()`.

Lines sort window centers within each contig, with breaks at missing values,
uncovered gaps and contig boundaries. A lone value appears as a point. Window
centers must be distinct within a contig. Circular lines interpolate genomic
position and value before drawing, so sparse points follow the ring; they do
not connect across the circle's origin. No smoothing or new GC estimates are
applied. Changing window width or step changes the summaries being plotted.

Bars span each interval and start at `baseline`, which defaults to the lower
limit. For a signed measurement, use e.g. `limits = c(-1, 1), baseline = 0`.
Missing bars leave gaps; missing heatmap values use `na.value`. Tracks with
only missing line/bar values retain the empty lane and axes, with no data key
in the legend. Bars and heatmaps draw overlapping intervals in input order.

You can also stack two heatmaps:

```{r multiple}
uncalled <- gene_gc
uncalled$value <- 100 * uncalled$n_ambiguous / (uncalled$end - uncalled$start)
p + syn_track(gene_gc, name = "GC (%)") +
  syn_track(uncalled, name = "Ambiguous bases (%)", palette = c("#FFF7EC", "#7F0000"))
```

`height` and `gap` are fractions of linear row spacing or the original circle
radius. Reduce them if many linear tracks need to fit between rows. Use
`limits` and `name` for other numeric measurements, for example
`syn_track(coverage, name = "Depth", limits = c(0, 200))`.

```{r scale}
p + syn_track(gene_gc) +
  scale_fill_syn_track(track = 1, palette = "casa_natal", breaks = c(0, 50, 100)) +
  ggplot2::theme(legend.position = "bottom")
```

The ordinary ggplot2 scale-replacement message is expected when replacing a
track scale. Numeric limits stay shared across all groups, making colors
comparable. `na.value` changes the color for explicitly missing measurements.

Add tracks before replacing coordinates. Faceting and coordinate transforms
are not supported for track placement. All three track types are static even
on interactive plots; existing gene and ribbon hover information still works.
Studio does not yet provide a track upload control.

## Design references

The arrangements were informed by [gbdraw](https://github.com/satoshikawato/gbdraw)
and the [circlize chord-diagram guide](https://jokergoo.github.io/circlize_book/book/the-chorddiagram-function.html).
This implementation uses [ggplot2's extension mechanism](https://ggplot2.tidyverse.org/articles/extending-ggplot2.html),
with polygon layers and a separate continuous aesthetic for every track.
Neither gbdraw nor circlize is a dependency. Existing plotting calls and the
32 ltc palette definitions are unchanged.
