---
title: "Circular synteny with ggplot2"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 8, collapse = TRUE,
                      comment = "#>", fig.align = "center")
library(ggsynteny)
```

The circular functions use the same tables as the linear functions and return
ordinary ggplot objects. Their arcs, curved arrows and ribbons are polygon
layers in Cartesian coordinates, with a fixed aspect ratio to keep the ring
circular. There is no graphics capture or external plotting backend.

## Chromosome-level synteny

```{r macro}
data(rice_sorghum)
p <- plot_circular_synteny(rice_sorghum, c("Rice", "Sorghum"),
                           palette = "casa_natal", chr_fill = "per_species")
p
```

Chromosome arc widths are proportional to chromosome lengths, using the same
scale across species. Ribbons attach at the supplied block coordinates. All
pairs among selected species are included, including within-species and
non-adjacent comparisons. Overlapping or repeated blocks remain separate
observations; their widths are not summed into a unique-coverage estimate.

`chr_fill` and `ribbon_fill` follow the linear function's coloring modes.
`gap = 1` separates chromosomes within species and `group_gap = 10` separates
species (degrees). `species_order` selects and orders genomes. Repeated
chromosome labels in different species are kept separate.

```{r customize}
p + ggplot2::labs(title = "Rice and sorghum",
                  caption = "Chromosome lengths inferred from MCScanX gene annotations") +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))
```

The default coverage ribbons join opposite interval boundaries for clarity.
When `blocks$orientation` is available, `show_orientation = TRUE` connects
genomic start/end coordinates according to `plus` or `minus`. For a plus
alignment, starts connect to starts; for a minus alignment, starts connect to
ends. Because both chromosomes may run clockwise around different portions
of a circle, a twist alone is not an inversion indicator. Read the orientation
metadata or interactive tooltip when interpreting alignment direction.
The previously bundled `rice_sorghum` object lacks orientation metadata;
newly parsed MCScanX input retains it.

## Gene-level synteny

```{r micro}
micro <- demo_microsynteny_data()
pm <- plot_circular_microsynteny(micro$features, micro$links,
                                palette = "casa_natal", ribbon_fill = "per_name")
pm
```

Contigs span the first supplied feature start through the last feature end.
Lengths share a scale across bins. This is an observed gene-region view;
unknown chromosome ends are not inferred. Each contig has its own coordinate
system and gap. The circular arrangement does not assert that an organism's
chromosomes are biologically circular. A feature crossing an origin needs to
be split into valid intervals before plotting.

Arrows show strand; ribbons represent homology. Gene strand alone is not used
to infer an alignment inversion. `ribbon_anchor = "body"` keeps arrowheads
clear. `ribbon_anchor = "full"` connects the entire gene span.

For the default identity coloring, the ramp spans 0 to 100 percent. Missing
identity values are gray. With no identity column, links use the same default
80-percent shade as the linear plot, without claiming a measured identity in
the tooltip. `palette` colors genes; set `ribbon_palette` explicitly to change
the identity ramp.

```{r identity}
plot_circular_microsynteny(micro$features, micro$links,
                           palette = "casa_natal", ribbon_palette = "Viridis")
```

## Inspect, export, and interact

The macro plot's `data` holds its sector layout. Genomic bounds remain in the
original unit; `theta_start` and `theta_end` are radians. Metadata attributes
record displayed links and features, so positions and filtering are inspectable.

```{r inspect}
p$data[, c("group_name", "sector_name", "start", "end")]
head(attr(pm, "circular_features")[, c("bin_id", "seq_id", "feat_id", "strand")])
```

```{r export, eval=FALSE}
ggplot2::ggsave("circular-synteny.pdf", p, width = 8, height = 8)
ggplot2::ggsave("circular-microsynteny.png", pm, width = 8, height = 8,
                dpi = 300, bg = "white")
```

```{r interactive, eval=requireNamespace("ggiraph", quietly=TRUE)}
pi <- plot_circular_microsynteny(micro$features, micro$links,
                                palette = "casa_natal", interactive = TRUE)
syn_girafe(pi, width_svg = 8, height_svg = 8)
```

The new functions introduce no dependencies. Optional interactivity uses the
same suggested ggiraph dependency as the linear views. Dense annotations can
be exported at a larger physical size or drawn with `label_genes = FALSE`;
labels are not silently discarded.

The visual references for this experiment are
[gbdraw's circular genome maps](https://github.com/satoshikawato/gbdraw/blob/main/docs/GALLERY.md)
and [circlize's chord diagrams](https://jokergoo.github.io/circlize_book/book/the-chorddiagram-function.html).
The latter uses relation strength for sector/link widths; ggsynteny's circular
functions instead retain genomic coordinate spans.
