---
title: "Circular synteny with ggplot2"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 8, collapse = TRUE,
                      comment = "#>", fig.align = "center")
library(ggsynteny)
```

The circular functions use the same tables as the linear functions and return
ordinary ggplot objects. Their arcs, curved arrows and ribbons are polygon
layers in Cartesian coordinates, with a fixed aspect ratio to keep the ring
circular. There is no graphics capture or external plotting backend.

## Chromosome-level synteny

```{r macro}
data(rice_sorghum)
p <- plot_circular_synteny(rice_sorghum, c("Rice", "Sorghum"),
                           palette = "casa_natal", chr_fill = "per_species")
p
```

Chromosome arc widths are proportional to chromosome lengths, using the same
scale across species. Ribbons attach at the supplied block coordinates. All
pairs among selected species are included, including within-species and
non-adjacent comparisons. Overlapping or repeated blocks remain separate
observations; their widths are not summed into a unique-coverage estimate.

`chr_fill` and `ribbon_fill` follow the linear function's coloring modes.
`gap = 1` separates chromosomes within species and `group_gap = 10` separates
species (degrees). `species_order` selects and orders genomes. Repeated
chromosome labels in different species are kept separate.

```{r customize}
p + ggplot2::labs(title = "Rice and sorghum",
                  caption = "Chromosome lengths inferred from MCScanX gene annotations") +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))
```

The default coverage ribbons join opposite interval boundaries for clarity.
When `blocks$orientation` is available, `show_orientation = TRUE` connects
genomic start/end coordinates according to `plus` or `minus`. For a plus
alignment, starts connect to starts; for a minus alignment, starts connect to
ends. Because both chromosomes may run clockwise around different portions
of a circle, a twist alone is not an inversion indicator. Read the orientation
metadata or interactive tooltip when interpreting alignment direction.
The previously bundled `rice_sorghum` object lacks orientation metadata;
newly parsed MCScanX input retains it.

## Gene-level synteny

```{r micro}
micro <- demo_microsynteny_data()
pm <- plot_circular_microsynteny(micro$features, micro$links,
                                palette = "casa_natal", ribbon_fill = "per_name")
pm
```

Contigs span the first supplied feature start through the last feature end.
Lengths share a scale across bins. This is an observed gene-region view;
unknown chromosome ends are not inferred. Each contig has its own coordinate
system and gap. The circular arrangement does not assert that an organism's
chromosomes are biologically circular. A feature crossing an origin needs to
be split into valid intervals before plotting.

Arrows show strand; ribbons represent homology. Gene strand alone is not used
to infer an alignment inversion. `ribbon_anchor = "body"` keeps arrowheads
clear. `ribbon_anchor = "full"` connects the entire gene span.

For the default identity coloring, the ramp spans 0 to 100 percent. Missing
identity values are gray. With no identity column, links use the same default
80-percent shade as the linear plot, without claiming a measured identity in
the tooltip. `palette` colors genes; set `ribbon_palette` explicitly to change
the identity ramp.

```{r identity}
plot_circular_microsynteny(micro$features, micro$links,
                           palette = "casa_natal", ribbon_palette = "Viridis")
```

## Three bacterial gene regions

The bundled bacterial inputs also provide a worked comparison of ZONMW-30,
ZONMW-20 and HI1: 21 genes on four contigs and nine supplied homology links.
These prepared tables can be used with either the circular or linear gene
view. The feature IDs are unique across all selected contigs.

```{r bacteria, fig.width=10, fig.height=10.5}
features <- read.delim(system.file("extdata", "circular_bacterial_features.tsv",
                                   package = "ggsynteny"))
links <- read.delim(system.file("extdata", "circular_bacterial_links.tsv",
                                package = "ggsynteny"))
pb <- plot_circular_microsynteny(
  features, links,
  bin_order = c("ZONMW-30", "ZONMW-20", "HI1"),
  palette = "casa_natal", ribbon_fill = "per_name",
  ribbon_alpha = 0.36, group_gap = 15, gap = 5,
  track_width = 0.055, label_size = 2.3, bin_label_size = 4.5
)
pb
```

The selection covers ZONMW-30 contig 4, ZONMW-20 contigs 10 and 17, and HI1
contig 1. Placeholder `z_spacer` rows are removed; retained gene coordinates
and intervening genomic gaps are unchanged. ZONMW-30's other contigs reuse
feature IDs and are outside this selected region. Each retained feature gets
a unique ID made from its bin, contig, coordinates and original ID; the latter
is retained in `source_feat_id`. Link endpoints must resolve to exactly one
feature, otherwise the preparation script stops.

Four input links connect ZONMW-30 to ZONMW-20, and five connect ZONMW-20 to
HI1. The input contains no direct ZONMW-30–HI1 links, and none are inferred.
It also has no identity scores, so ribbons are coloured by gene name. This
is a comparison of selected gene regions, not a whole-genome alignment.

From a source checkout, `Rscript data-raw/circular_bacteria.R` rebuilds the
prepared TSVs, the README PNG and the annotated vector PDF. The original
`bacterial_genes.csv` and `bacterial_links.csv` inputs are preserved.

```{r bacteria-export, eval=FALSE}
ggplot2::ggsave("three-bacteria.pdf", pb, width = 10, height = 10.5)
```

## Inspect, export, and interact

The macro plot's `data` holds its sector layout. Genomic bounds remain in the
original unit; `theta_start` and `theta_end` are radians. Metadata attributes
record displayed links and features, so positions and filtering are inspectable.

```{r inspect}
p$data[, c("group_name", "sector_name", "start", "end")]
head(attr(pm, "circular_features")[, c("bin_id", "seq_id", "feat_id", "strand")])
```

```{r export, eval=FALSE}
ggplot2::ggsave("circular-synteny.pdf", p, width = 8, height = 8)
ggplot2::ggsave("circular-microsynteny.png", pm, width = 8, height = 8,
                dpi = 300, bg = "white")
```

```{r interactive, eval=requireNamespace("ggiraph", quietly=TRUE)}
pi <- plot_circular_microsynteny(micro$features, micro$links,
                                palette = "casa_natal", interactive = TRUE)
syn_girafe(pi, width_svg = 8, height_svg = 8)
```

The new functions introduce no dependencies. Optional interactivity uses the
same suggested ggiraph dependency as the linear views. Dense annotations can
be exported at a larger physical size or drawn with `label_genes = FALSE`;
labels are not silently discarded.

The visual references for this experiment are
[gbdraw's circular genome maps](https://github.com/satoshikawato/gbdraw/blob/main/docs/GALLERY.md)
and [circlize's chord diagrams](https://jokergoo.github.io/circlize_book/book/the-chorddiagram-function.html).
The latter uses relation strength for sector/link widths; ggsynteny's circular
functions instead retain genomic coordinate spans.
