---
title: "Interactive synteny plots"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  out.width = "100%"
)
```

Every ggsynteny plot can be made interactive with
[ggiraph](https://davidgohel.github.io/ggiraph/): build the plot with
`interactive = TRUE`, then render it with `syn_girafe()`. Hovering a
ribbon fades everything else and shows a tooltip with the block (or link)
coordinates; hovering a chromosome or gene shows what it is. The widgets
work anywhere HTML does — R Markdown, Quarto, Shiny, pkgdown.

```{r setup}
library(ggsynteny)
```

## Macro-synteny: hover the rice-sorghum blocks

*Hover a ribbon* to isolate one collinear block and read its coordinates;
*hover a chromosome* for its size. Scroll/pinch to zoom (up to 4x).

```{r interactive-macro}
data(rice_sorghum)

p <- plot_synteny(rice_sorghum, c("Rice", "Sorghum"),
                  palette = "casa_natal",
                  chr_fill = "per_chr", ribbon_fill = "source_chr",
                  interactive = TRUE)
syn_girafe(p, width_svg = 10, height_svg = 5)
```

## Micro-synteny: hover genes and homology links

Gene tooltips show name, contig, coordinates and strand; link tooltips show
the connected gene pair and its percent identity:

```{r interactive-micro}
micro <- demo_microsynteny_data()

p <- plot_microsynteny(micro$features, micro$links,
                       bin_order = c("ZONMW-30", "ZONMW-20", "ZONMW-10"),
                       palette = "casa_natal",
                       interactive = TRUE)
syn_girafe(p, width_svg = 10, height_svg = 4.5)
```

## Styling the widget

`syn_girafe()` exposes the SVG size, the hover CSS, and takes any extra
`ggiraph::opts_*()` through `opts`:

```{r interactive-styled, eval = FALSE}
syn_girafe(p,
           width_svg = 12, height_svg = 5,
           hover_css = "fill-opacity:1;stroke:#000;stroke-width:1.5px;",
           opts = list(ggiraph::opts_toolbar(position = "topright")))
```

Everything else about the plot — palettes, fills, curvature — is exactly
as in the static version, so `interactive = TRUE` can be flipped on at the
very end of your workflow.
