---
title: "Real data: rice-sorghum and a bacterial gene cluster"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 10, fig.height = 5.5, dpi = 150,
  out.width = "100%"
)
```

This article works through two *real* datasets end to end — one
macro-synteny (chromosome-level) and one micro-synteny (gene-level) — the
way you would use ggsynteny on your own data.

```{r setup}
library(ggsynteny)
library(ggplot2)
```

## Macro-synteny: rice vs sorghum, from MCScanX output

Rice (*Oryza sativa*) and sorghum (*Sorghum bicolor*) diverged roughly 50
million years ago, yet their chromosomes still align over long stretches —
the textbook example of grass synteny.

The bundled `rice_sorghum` dataset is genuine MCScanX output: the
rice-sorghum example data shipped with MCScanX itself (all-vs-all protein
BLAST plus gene positions; Wang et al. 2012, *Nucleic Acids Research*
40:e49) was run through MCScanX with default parameters, and the resulting
`os_sb.collinearity` file was parsed with ggsynteny's own
`read_mcscanx()`:

```r
syn <- read_mcscanx("os_sb.collinearity", "os_sb.gff")
```

The prepared object keeps the 100 largest inter-genome blocks (at least 20
collinear gene pairs each) so the default plot stays legible —
`data-raw/rice_sorghum.R` in the package sources documents every step:

```{r rice-sorghum-data}
data(rice_sorghum)
str(rice_sorghum, max.level = 1)
head(rice_sorghum$blocks, 3)
```

One call plots it. Colouring the ribbons by their rice source chromosome
makes the block structure pop — note how rice 1 maps almost entirely to
sorghum 3, and rice 11/12 to sorghum 5/8:

```{r rice-sorghum-plot, fig.height = 6}
plot_synteny(rice_sorghum, c("Rice", "Sorghum"),
             palette = "casa_natal",
             chr_fill = "per_chr", ribbon_fill = "source_chr")
```

The `blocks` data frame keeps MCScanX's alignment score and gene count per
block, so you can filter harder before plotting:

```{r rice-sorghum-filtered, fig.height = 6}
big <- rice_sorghum
big$blocks <- subset(big$blocks, n_genes >= 100)

plot_synteny(big, c("Rice", "Sorghum"),
             palette = "minou", ribbon_alpha = 0.45)
```

## Micro-synteny: a bacterial moa/moe gene cluster

The package also ships a gene-level dataset from a real comparative
analysis: the molybdenum-cofactor (moa/moe) gene cluster across three
bacterial strains, with per-link protein identities. The raw CSVs live in
`inst/extdata`, so this doubles as a template for loading your own files:

```{r bacterial-data}
genes <- read.csv(system.file("extdata", "bacterial_genes.csv",
                              package = "ggsynteny"))
links <- read.csv(system.file("extdata", "bacterial_links.csv",
                              package = "ggsynteny"))

head(genes, 3)
head(links, 3)
```

The links file pairs `feat_id`s across strains; rename to the columns
`plot_microsynteny()` expects:

```{r bacterial-links}
links <- data.frame(feat_id_a = links$feat_id,
                    feat_id_b = links$feat_id2)
```

```{r bacterial-plot, fig.height = 5}
plot_microsynteny(genes, links,
                  bin_order = c("ZONMW-30", "ZONMW-20", "ZONMW-10"),
                  palette = "casa_natal")
```

Without an `identity` column the ribbons fall back to a uniform colour via
`ribbon_fill = "identity"`'s default; with identities (as in
`demo_microsynteny_data()`) the ribbon darkness encodes percent identity,
and an ordered palette restyles the ramp:

```{r bacterial-identity, fig.height = 5}
micro <- demo_microsynteny_data()

plot_microsynteny(micro$features, micro$links,
                  bin_order = c("ZONMW-30", "ZONMW-20", "ZONMW-10"),
                  gene_fill = "per_name", gene_palette = "casa_natal",
                  ribbon_fill = "identity", ribbon_palette = "heatmap0")
```

## References

- Wang, Y. et al. (2012). "MCScanX: a toolkit for detection and
  evolutionary analysis of gene synteny and collinearity."
  *Nucleic Acids Research* 40(7):e49. <doi:10.1093/nar/gkr1293>
